---
title: "Revisiting Selecting Medical Hardware Using Pairwise Comparisons"
author: "Tim Anderson"
format: pdf
editor: visual
---

# Cochlear Implant Decision Making

## Introduction

In 2016, [Anderson and Jahromi](https://pdxscholar.library.pdx.edu/etm_fac/93/) demonstrated medical device decision making of a cochlear implant system from a patient's perspective using hierarchical decision making. This paper returns to this application demonstrating two different but related pair-wise comparison decision making techniquesanalytical hierarchy process, AHP, and Hierarchical Decision Modeling, HDM.

Furthermore, it is updated to reflect currently available hardware and evolving user needs and compares the results to the 2016 paper.

The model is also simplified and the analysis is done using packages in R as a Quarto markdown document.

## Background

The earlier PICMET paper used 8 characteristics divided into three categories:

-   Performance
    -   Current Implant Technology
    -   Main External Processor Unit
    -   Main External Unit's Ingress Protection (IP Rating)
-   Accessories
    -   Secondary External Process Unit
    -   Connection Options
    -   Bonuses
-   Future Issues
    -   Future Implant Capability
    -   Warranty and Reliability

## Revised Model

This paper simplifies the model to just four characteristics of a cochlear implant system.

-   **Implant:** Technology for internally implanted device's current capabilities. While anecdotal comments and studies do not appear to find clear differences in success outcomes by manufacturers, the technologies are different.
-   **Flexibility:** Variety of benefits for the device, including flexibility such as off-the-ear versus on--the-ear for an alternative external processor. It is important to use the device and throughout the day to help the brain adapt to the new way of hearing. For users with eyeglasses and mask wearing, an off the ear processor option can be a very useful option.
-   **Connectivity:** This reflects the general connectivity capabilities of the current processor(s) such as direct Bluetooth with requiring additional devices, and Auracast (a newly announced multipoint Bluetooth capability).
-   **Future:** Anticipated future prospects both in terms of company commitment and the opportunity for future processors to leverage more capabilities of the implant.

Some other factors to consider might include comfort for wearing, battery life, frequency of recalls, commitment to support out of warranty products, support community, familiarity of clinic/surgeon with the manufacturer.

## Analysis

Let's start by loading two packages. The first, `ahpsurvey`, is for doing Analytic Hierarchy Process and is designed with an emphasis on incorporating many decision makers. The second package, `kableExtra`, is for doing better formatted tables.

```{r}
library (ahpsurvey)
library (kableExtra)
```

First we will construct the data comparing the relative attributes.

```{r}
atts <- c("impl", "flex", "conn", "fut")

matA <- matrix(c(-2,-2,-2,1,2,2), ncol=6)

colnames (matA) <- c("imp_flex", "imp_conn", "impl_fut", 
                     "flex_conn", "flex_fut", 
                     "conn_fut")

kbl(matA , booktabs=T, 
    caption="Patient's preferences of characteristics.") |> 
kableExtra::kable_styling(latex_options =  "hold_position")  
```

The `ahp` command takes a matrix, in this case `matA`, of decision maker information describing the relative importance for the decision maker of each characteristic. For example, the value of `conn_fut` reflects a comparison of Connectivity to Future Capabilities of 2 indicates that

```{r}
matAr <- ahpsurvey::ahp(matA, 
                          atts=atts, 
                          negconvert=FALSE, # Default of FALSE
                          reciprocal=TRUE, # Defauult of TRUE
                          method="eigen", # Default of Eigenvalue
                          aggmethod="geometric", # Default Geometric mean
                          )
kbl (matAr, booktabs=T, 
     caption="Patient's preferences of characteristics.") |> 
kableExtra::kable_styling(latex_options =  "hold_position")      
```

Just as there was in 2016, there are three United States FDA approved cochlear implant manufacturers: Advanced Bionics, Cochlear, and Med-El. The result is that for US patients, it typically is a choice between just these three manufacturers although sometimes one or two manufactures are not options due to patient's unique clinical characteristics, insurance coverage, or clinic limitations.

The patient's decision is then two-fold, first whether to get a cochlear implant and secondly which manufacturer to select. A third decision could be which device to select since each manufacturer offers a standard behind-the-ear (BTE) device processor and then a secondary off-the-ear (OTE) processor option.

The processor devices are summarized below

-   Advanced Bionics

    -   BTE: **Naida M90** was approved by the FDA in 2020 and is the only cochlear implant that fully supports Bluetooth phone calls, including the speaker's voice.

    -   OTE: **Neptune**, a processor tailored around swimming.

-   Cochlear

    -   BTE: **Nucleus 7** was FDA approved in 2017 but the Nucleus 8 has been announced outside the US and US availability is pending FDA approval, expected in late 2022 or early 2023. The **Nucleus 8** is expected to have broad Bluetooth protocol support including the new Auracast protocol but will still not have the capability to relay the wearer's voice back to the telephone or computer.

    -   OTE: **Kanso 2** was FDA approved in 2020 and is a single disc processor atop the magnet to avoid interfering with eyeglasses and facemasks.

-   Med-El

    -   BTE: **Sonnet 2** was FDA approved in 2020 and can support Bluetooth connections with a battery attachment.

    -   OTE: **Rondo 3** was also FDA approved in 2020. The Rondo 3 does not have any direct Bluetooth capability on its own and requires a separate device from Med-El.

The following section represents the user's perceived strengths of the three manufacturers with respect to each of the four characteristics. Note that these are all based on a patient's current perspective based on a wide-ranging exploration but do not represent medical professional's judgement. Furthermore, these perceived benefits are based on the patient's anticipated usage. Just as the patient's preferences for the four characteristics should not be interpreted as generalizing to other patients, the relative strengths of cochlear implant manufacturers on each of these characteristics might vary for other other patients.

Let's start by examining Implant Technology. In this case we will assign a value of 4 for the relative strength of Advanced Bionics' implant technology to that of Cochlear's. A 4 is between "slightly more important" and "moderately more important." After populating the vector (one-dimensional matrix) of `matB1`, we then place the result in `matB1r`.

```{r}

alts <- c("AB", "Coch", "Med")   # Device Alteratives

# Implant Technology Pairwise Comparisons
matB1 <- matrix(c(6, 1, -4), ncol=3)
#colnames(matB1) <- c("AB_Coch", "AB_Med", "Coch_Med")

# Analysis by AHP  
matB1r <- ahpsurvey::ahp(matB1, atts=alts)
       
```

Recall that in this case a negative value in a pairwise comparison reflects what Saaty's AHP would use as a a reciprocal of the absolute value. For example, assume that a device A compared to device B had a -5 with respect to a characteristic. This means that B is strongly preferred on this characteristic to A. In Saaty's notation, this would be denoted as as $\frac{1}{5}$ instead of -5. The negative value is used to avoid issues of displaying significant digits.

```{r echo=FALSE}

knitr::kable (matB1r, booktabs=T,
              caption="Implant Technology")|> 
  kableExtra::kable_styling(latex_options =  "hold_position")  
```

In this case we can see that the Implant Technology strongly favors Advanced Bionics followed by Med-El.

```{r}
# Flexibility 
matB2 <- matrix(c(-7, -5, 1), ncol=3)
  c("AB_Coch", "AB_Med", "Coch_Med")
matB2r <- ahpsurvey::ahp(matB2, atts=alts)

# Connectivity:  Current Connectivity options and support 
matB3 <- matrix(c(2, 3, 2), ncol=3)
  c("AB_Coch", "AB_Med", "Coch_Med")
matB3r <- ahpsurvey::ahp(matB3, atts=alts)

# Future:  Future expansion capability and company prospects 
matB4 <- matrix(c(6, 1, -5), ncol=3)
  c("AB_Coch", "AB_Med", "Coch_Med")
matB4r <- ahpsurvey::ahp(matB4, atts=alts)
```

## Integrating Characteristic Preferences and Alternatives

Now it is time to put everything together.

Recall the collective judgments for how Cochlear Implant Systems compare on each characteristic.

```{r}
judgments <- rbind (matB1, matB2, matB3, matB4)
rownames (judgments)<-c("Implant Technology", "Flexibility", 
                        "Connectivity", "Future")
colnames (judgments)<-   c("AB_Coch", "AB_Med", "Coch_Med")
knitr::kable (judgments, booktabs=T,
              caption="Relative judgements of systems on each characteristic")|> 
kableExtra::kable_styling(latex_options =  "hold_position")  
```

As we saw earlier, each of the analysis results has three components, the integrated results, the consistency ratio, and the relative ranking. For numerical purposes, we will extract the scores. We'll start with the weights on the characteristics.

```{r}
char_weights <- matAr[1:4]

knitr::kable (char_weights, booktabs=T,
caption="Weights for each characteristic")|> 
kableExtra::kable_styling(latex_options =  "hold_position")  
```

```{r echo=FALSE}
# ---------------------
# Displaying results
   
knitr::kable (matB2r, booktabs=T,
caption="Flexibility of System: For Example Off the Ear Processor Option")|> 
kableExtra::kable_styling(latex_options =  "hold_position")            

knitr::kable (matB3r, booktabs=T,
caption="Connectivity:  Comprehensiveness and Robustness of Connection Options such as Bluetooth, Auracast, Speaking on Phone Calls")|> 
kableExtra::kable_styling(latex_options =  "hold_position")            

knitr::kable (matB4r, booktabs=T,
caption="Future: Long-term Prospects for Improved Capabilities")|> 
kableExtra::kable_styling(latex_options =  "hold_position")           
```

Now, let's do the same thing for each of the device characteristics.

```{r echo=FALSE}
device_weights <- rbind (matB1r[1:3], matB2r[1:3], matB3r[1:3], matB4r[1:3])

rownames(device_weights)<-c("Implant Technology","Flexibility",
                            "Connectivity", "Future")

knitr::kable (device_weights, booktabs=T,
caption="Weights for each device on each characteristic")|> 
kableExtra::kable_styling(latex_options =  "hold_position")  
```

The last step is to simply multiply the device weights and characteristic weights to come up with an overall score.

```{r}
final_score <- t(device_weights) %*% t(char_weights)

```

```{r echo=FALSE}
knitr::kable (final_score, booktabs=T,
caption="Final score for each manufacturer.")|> 
kableExtra::kable_styling(latex_options =  "hold_position") 
```

In this case, it appears that the preferred alternative is Advanced Bionics.

The patient has until December to make a final decision and will continue to revise the model and results with additional information.

## HDM Analysis

Hierarchical Decision Modeling, HDM, similar to AHP but has been widely employed in the Engineering and Technology Management Department.

### Loading R Script for HDM

Kevin van Blommestein wrote an R script for doing HDM. Let's load this in script to make it available for the analysis.

## Three Factor Model

```{r, include=FALSE}
source("HDM.R", local = knitr::knit_global())
# or sys.source("your-script.R", envir = knitr::knit_global())
```

```{r, echo=FALSE, fig.cap = "Cochlear Implant Device Selection"}
library(DiagrammeR)
library (webshot2)
mermaid("
graph TD
  A(Cochlear Implant Company Selection)-->B(Implant)
  A-->C(Connectivity)
  A-->D(Future)
  B-->J(Advanced Bionics)
  B-->K(Cochlear LLC)
  B-->M(MedEl)
  C-->J
  C-->K
  C-->M
  D-->J
  D-->K
  D-->M
", height = '100%', width = '100%')
```

### Conducting HDM

```{r}
# A, B, and C are characteristics
# A:  Convenience of Processor
# B:  Electrode Technology
# C:  Connectivity

d1 <- data.frame(A_B = c(90), A_C = c(90), B_C = c(90))
# Name rows with unique expert names
row.names(d1) <- c("User")

m <- pwc.matrix(d1, one.side = T)

pc <- pwc.weights(m)

pc$detail$weights
```

### Interpretation of HDM Results

Repeat for each

Five Factor HDM Model

```{r}
library(DiagrammeR)
mermaid("
graph TD
  A(Cochlear Implant Company Selection)-->B(Implant)
  A-->C(Flexibility)
  A-->D(Connectivity)
  A-->E(Community)
  A-->F(Clinic)
  B-->J(Advanced Bionics)
  B-->K(Cochlear LLC)
  B-->M(MedEl)
  C-->J
  C-->K
  C-->M
  D-->J
  D-->K
  D-->M
  E-->J
  E-->K
  E-->M
  F-->J
  F-->K
  F-->M
")
```

```{r}
# A, B, C, D, and E are characteristics
# A:  Implant
# B:  Flexibility
# C:  Connectivity
# D:  Company
# E:  Clinic

# A_B=35 indicates that A is given 65 and B is given 35 points.
#     In other words, A is much more important than B.

d2_0 <- data.frame(A_B=c(45), A_C=c(30), A_D=c(30), A_E=c(40),
                              B_C=c(40), B_D=c(40), B_E=c(50),
                                         C_D=c(50), C_E=c(60),
                                                    D_E=c(60))
# Name rows with unique expert names
row.names(d2_0) <- c("Characteristic")

m2_0 <- pwc.matrix(d2_0, one.side = T)

pc2_0 <- pwc.weights(m2_0)

pc2_0$detail$weights
```

Now let's examine each of the options with respect the features.

```{r}
# A, B, and C are alernatives
# A:  Advanced Bionics
# B:  Cochlear
# C:  Med-El

# A_B=20 indicates that A (Adv. Bionics) gets 80 points, Cochlear gets 20
#    on this characteristic.

d2_1 <- data.frame(A_B=c(20), A_C=c(53),    # Implant
                              B_C=c(82))
row.names(d2_1) <- c("Implant")

m2_1 <- pwc.matrix(d2_1, one.side = T)
pc2_1 <- pwc.weights(m2_1)

pc2_1$detail$weights
```

```{r}
d2_2 <- data.frame(A_B=c(90), A_C=c(90),    # Flexibility
                              B_C=c(50))
row.names(d2_2) <- c("Flexibility")

m2_2 <- pwc.matrix(d2_2, one.side = T)
pc2_2 <- pwc.weights(m2_2)

pc2_2$detail$weights
```


```{r}
d2_3 <- data.frame(A_B=c(45), A_C=c(10),    # Connectivity
                              B_C=c(12))
row.names(d2_3) <- c("Connectivity")

m2_3 <- pwc.matrix(d2_3, one.side = T)
pc2_3 <- pwc.weights(m2_3)

pc2_3$detail$weights
```

```{r}
d2_4 <- data.frame(A_B=c(60), A_C=c(65),    # Community
                              B_C=c(53))
row.names(d2_4) <- c("Community")

m2_4 <- pwc.matrix(d2_4, one.side = T)
pc2_4 <- pwc.weights(m2_4)

pc2_4$detail$weights
```



```{r}
d2_5 <- data.frame(A_B=c(30), A_C=c(50),    # Clinic
                              B_C=c(70))
row.names(d2_5) <- c("Clinic")

m2_5 <- pwc.matrix(d2_5, one.side = T)
pc2_5 <- pwc.weights(m2_5)

pc2_5$detail$weights
```

```{r}
alts2 <- rbind(pc2_1$detail$weights,
              pc2_2$detail$weights,
              pc2_3$detail$weights,
              pc2_4$detail$weights,
              pc2_5$detail$weights)

res2 <- t(alts2)%*%t(pc2_0$detail$weights)

incon2 <- rbind(pc2_1$summary$inconsistency, 
               pc2_2$summary$inconsistency, 
               pc2_3$summary$inconsistency, 
               pc2_4$summary$inconsistency, 
               pc2_5$summary$inconsistency) 
```


Let's summarize the results in a few tables.

```{r}
kbl(cbind(alts2, incon2, t(pc2_0$detail$weights)), 
    booktabs=T, digits=5,
    caption="Weights by Alternative",
    col.names = c("Adv. Bionics (A)","Cochlear (B)", "Med-El (C)", 
                  "Inconsistency",
                  "Importance"))
```

These results show a number of interesting things:

We can see that Med-El generally well on each characteristic other than connectivity.

- Each of the pair-wise comparison matrices for characteristics had very low values of inconsistency suggesting that the decision maker is at least internally consistent.
- The importance of the characteristics varies.  The implant technology is permanent whereas the external processors can change over time with regular upgrades.  Not surprisingly, the implant then gets the highest weight.


Now what we need to do is to combine the results by multiplying the performance on each characteristic by the importance.


```{r}
res2 <- t(alts2)%*%t(pc2_0$detail$weights)
kbl(t(res2), 
    booktabs=T, digits=5,
    caption="Overall Score for Five Characteristic HDM Model",
    col.names = c("Adv. Bionics (A)","Cochlear (B)", "Med-El (C)"))
```


## Future Work

This work could be extended in several directions.

-   Refining the model in terms of characteristics.

-   Applying HDM in place of AHP

-   Testing the impact of AHP alternatives

-   Verifying the model with other patients

-   Gathering responses from other patients
